The relative roles of genetic, ecological and environmental factors in insular species radiations

### 2018-05-14 paper/draft.tex
begin restructuring intro

%% major topic: besides classical mechanisms (ecological, environmental), genetic mechanisms affect speciation/radiations

%% Islands
Islands are understood as suitable model systems for ecology and evolution due to their well definable characteristics
like isolation (e.g. Losos \& Ricklefs 2009; Warren et al. 2015  %%Islands as model systems in ecology and evolution: prospects fifty years after MacArthur‐Wilson. EcoLett). %TODO: ref
Part of this value as study system originates by typically remarkable species.
In fact, islands often possess unique species communities and can exceede mainland areas
by a factor of 9.5 and 8.1 in plant and animal  endemism, respectively
(Kier et al. 2005. %A global assessment of endemism and species richness across island and mainland regions. PNAS). %TODO: refs
Such high endemism makes islands  especially threatened by human impacts, in particular climate change and invasive species (Fordham \& Brook 2010). %TODO: refs
Hence, understanding the origin, maintenance, functioning and evolution of island communities is of a high priority objective.

For the prediction of island community size, several theoretical frameworks have been proposed,
including one of the first and to this day still highly influential classical ecological theories,
namely MacArthur and Wilson's equilibrium theory of island biogeography (ETIB). %TODO: ref
In it, species richness on islands is the result of opposing rates of immigration and extinction,
both of which in turn are modified by island size (area) and its distance from the mainland (isolation).
This relationship was since successfully applied to a multitude of islands (Lomolina \& Weiser 2001, Triantis et al.\ 2011)%TODO: refs
, but also island like systems, such as coral reefs, lakes/ponds or forest fragments (Kelly et al.\ 1989, Peay et al.\ 2007, ). %TODO: refs
This general applicability (Lomolino 2000), however, cannot hide the fact
that the ETIB fails to formalize a mechanistic explanation of how extinction and immigration are modified by area and isolation, respectively.
Additionally, it has been demonstrated more recently that area also affects immigration via target effects (REFs) a
nd isolation affects extinction via rescue-effects (REFs).
Finally, ETIB has particularly been criticized for its lack of considering speciation (see e.g.\ Losos \& Schluter 2000). %TODO: refs

About a decade ago, island biogeography theory gained new momentum by Whittaker et al.'s General Dynamic Model (GDM, Whittaker et al.\ 2007, 2008). %TODO: ref
The conceptual framework traces island communities along the geomorphological trajectory of hotspot islands,
whith islands emerging from sea level via volcanic hotspot activity,
increasing over time in size and topographic complexity,
then eroding and subsiding in the ocean again when the island is not over the hotspot anymore.
In doing so, the model incorporates concepts of niche theory, speciation and species succession.
Yet, there are still many processes and drivers deemed important,
which are not considered by any formal theory and often little investigated
%[if you say this, you have to mention at leastr a couple os examples, otherwise it is an empty claim].

%% Models
The problem of investigating these processes is the time frame and study extent needed for a proper evaluation.
Because islands biogeography takes place at evolutionary time frames (millions of years),
comprehensive and meaningful data is difficult and virtually impossible to obtain.
Therefore, there has been an interest in building predictive models for islands.
In a recent review, we indeed found a rapid increase in number of mechanistic simulaiton models for island biogeography %\cite{leidinger2017biodiversity}
(Leidinger \& Sarmento Cabral 2017).
Although many of the reviewed models integrate multiple processes at a time, unfortunately,
most models failed to include microevolutionary or more explicit genetic processes %\cite{leidinger2017biodiversity}
(Leidinger \& Sarmento Cabral 2017).
One direction towards this is given by models using concepts such as ``protracted speciaton'' (Rosindell \& Harmon 2011) %\cite{rosindellXXX},
where speciation occurs with a delay from the actual
divergence event if the population remains isolated during that time.
But nature is usually more complex than this (cf.\ Silvertown 2004, but see Herben et al.\ 2005)
and genetic mechanisms should also play a central role for speciation/radiation and species richness.
Questions, such as "[cite two questions from Patino et al. 2017]", thus remain largely open (Patiño et al.\ 2017)
and may require explicit consideration of evolutionary processes.
%% Yet, there are numerous examples for fascinating patterns of trait distributions and evolution on islands, such as island syndromes[, ...] or the fact of explosively radiating lineages (Hughes \& Eastwood 2006, Kocher 2004)
%%, while others remain one single species (Price/Chase) (eg. \cite{price2004speciation}).
%% Also: Ecological Opportunity and Adaptive Radiation, Stroud \& Losos 2016: ``Does the concept have predictive value[...]?'' \cite{stroud2016ecological}
%% Phylogenetic niche conservatism, phylogenetic signal and the relationship between phylogenetic relatedness and ecological similarity among species, Losos 2008: ecological vs. phylogenetic similarity \cite{losos2008phylogenetic}
Still, the most mechanisms integrated in island models [?] so far include only ecological
or geo-morphological processes (Rundell \& Price 2009, Esselstyn et al. 2011),
while genetic properties are hardly recognized in the island context.
This is surprising given the essential role islands have played in the development of evolutionary theories.
Nevertheless, an increasing number of studies investigates genetic processes and
their effect on reproductive/ecological isolation of populations (e.g., Via \& West 2008, Via 2001 and references therein)
as well as evolutionary rescue provided by the feedback with ecological processes (Schiffers et al. 2013).
%%TODO: here come a list of processes with examples
%% - genetic processes in general
%% - isolation processes
Besides arthropods (Howthorne \& Via 2001), these processes can be observed in plants as well. %Are you talking about islands here? Be clearer, because you said about the lack of geentic studies in island contexts.
For example, the rapid switching of pollinator syndromes in species of the genus \textit{Petunia} %use package for this!
seems to be a consequence of the tight genetic linkage between loci for pollination syndrome traits (Hermann et al.\ 2013). % cite properly!
Considering these phenomena, we expect speciation rates to be highest at intermediate levels of gene linkage --- low
levels leads to dissolving phenotype populations, whereas high levels hinder rapid adaption. %what Schiffers et al say about linkage and evolutionary rescue (basically the opposite of diversification?)

>>>>

%% major topic: besides classical mechanisms (ecological, environmental), genetic mechanisms affect speciation/radiations
Island radiations like the Darwin finches on Galápagos or \textit{Aeonium} on the Canaries have fascinated researchers throughout history %% cite? Darwin? ...?
To this date, however, it is unclear why these genera produced this multitude of species as opposed to most other species groups.
%%
Nevertheless, an increasing number of studies investigates genetic processes and
their effect on reproductive/ecological isolation of populations (e.g., Via \& West 2008, Via 2001 and references therein)
as well as evolutionary rescue provided by the feedback with ecological processes (Schiffers et al. 2013).
%%TODO: here come a list of processes with examples
%% - genetic processes in general
%% - isolation processes
Besides arthropods (Howthorne \& Via 2001), these processes can be observed in plants as well. %Are you talking about islands here? Be clearer, because you said about the lack of geentic studies in island contexts.
For example, the rapid switching of pollinator syndromes in species of the genus \textit{Petunia} %use package for this!
seems to be a consequence of the tight genetic linkage between loci for pollination syndrome traits (Hermann et al.\ 2013). % cite properly!
Considering these phenomena, we expect speciation rates to be highest at intermediate levels of gene linkage --- low
levels leads to dissolving phenotype populations, whereas high levels hinder rapid adaption. %what Schiffers et al say about linkage and evolutionary rescue (basically the opposite of diversification?)

%% Islands
Islands are understood as suitable model systems for ecology and evolution due to their well definable characteristics
like isolation (e.g. Losos \& Ricklefs 2009; Warren et al. 2015  %%Islands as model systems in ecology and evolution: prospects fifty years after MacArthur‐Wilson. EcoLett). %TODO: ref
Part of this value as study system originates by typically remarkable species.
In fact, islands often possess unique species communities and can exceede mainland areas
by a factor of 9.5 and 8.1 in plant and animal  endemism, respectively
(Kier et al. 2005. %A global assessment of endemism and species richness across island and mainland regions. PNAS). %TODO: refs
Such high endemism makes islands  especially threatened by human impacts, in particular climate change and invasive species (Fordham \& Brook 2010). %TODO: refs
Hence, understanding the origin, maintenance, functioning and evolution of island communities is of a high priority objective.
For the prediction of island community size, several theoretical frameworks have been proposed,
including one of the first and to this day still highly influential classical ecological theories,
namely MacArthur and Wilson's equilibrium theory of island biogeography (ETIB). %TODO: ref
In it, species richness on islands is the result of opposing rates of immigration and extinction,
both of which in turn are modified by island size (area) and its distance from the mainland (isolation).
This relationship was since successfully applied to a multitude of islands (Lomolina \& Weiser 2001, Triantis et al.\ 2011)%TODO: refs
, but also island like systems, such as coral reefs, lakes/ponds or forest fragments (Kelly et al.\ 1989, Peay et al.\ 2007, ). %TODO: refs
This general applicability (Lomolino 2000), however, cannot hide the fact
that the ETIB fails to formalize a mechanistic explanation of how extinction and immigration are modified by area and isolation, respectively.
Additionally, it has been demonstrated more recently that area also affects immigration via target effects (REFs) a
nd isolation affects extinction via rescue-effects (REFs).
Finally, ETIB has particularly been criticized for its lack of considering speciation (see e.g.\ Losos \& Schluter 2000). %TODO: refs
About a decade ago, island biogeography theory gained new momentum by Whittaker et al.'s General Dynamic Model (GDM, Whittaker et al.\ 2007, 2008). %TODO: ref
The conceptual framework traces island communities along the geomorphological trajectory of hotspot islands,
whith islands emerging from sea level via volcanic hotspot activity,
increasing over time in size and topographic complexity,
then eroding and subsiding in the ocean again when the island is not over the hotspot anymore.
In doing so, the model incorporates concepts of niche theory, speciation and species succession.
Yet, there are still many processes and drivers deemed important,
which are not considered by any formal theory and often little investigated
%[if you say this, you have to mention at leastr a couple os examples, otherwise it is an empty claim].

%% Models
The problem of investigating these processes is the time frame and study extent needed for a proper evaluation.
Because islands biogeography takes place at evolutionary time frames (millions of years),
comprehensive and meaningful data is difficult and virtually impossible to obtain.
Therefore, there has been an interest in building predictive models for islands.
In a recent review, we indeed found a rapid increase in number of mechanistic simulaiton models for island biogeography %\cite{leidinger2017biodiversity}
(Leidinger \& Sarmento Cabral 2017).
Although many of the reviewed models integrate multiple processes at a time, unfortunately,
most models failed to include microevolutionary or more explicit genetic processes %\cite{leidinger2017biodiversity}
(Leidinger \& Sarmento Cabral 2017).
One direction towards this is given by models using concepts such as ``protracted speciaton'' (Rosindell \& Harmon 2011) %\cite{rosindellXXX},
where speciation occurs with a delay from the actual
divergence event if the population remains isolated during that time.
But nature is usually more complex than this (cf.\ Silvertown 2004, but see Herben et al.\ 2005)
and genetic mechanisms should also play a central role for speciation/radiation and species richness.
Questions, such as "[cite two questions from Patino et al. 2017]", thus remain largely open (Patiño et al.\ 2017)
and may require explicit consideration of evolutionary processes.
%% Yet, there are numerous examples for fascinating patterns of trait distributions and evolution on islands, such as island syndromes[, ...] or the fact of explosively radiating lineages (Hughes \& Eastwood 2006, Kocher 2004)
%%, while others remain one single species (Price/Chase) (eg. \cite{price2004speciation}).
%% Also: Ecological Opportunity and Adaptive Radiation, Stroud \& Losos 2016: ``Does the concept have predictive value[...]?'' \cite{stroud2016ecological}
%% Phylogenetic niche conservatism, phylogenetic signal and the relationship between phylogenetic relatedness and ecological similarity among species, Losos 2008: ecological vs. phylogenetic similarity \cite{losos2008phylogenetic}
Still, the most mechanisms integrated in island models [?] so far include only ecological
or geo-morphological processes (Rundell \& Price 2009, Esselstyn et al. 2011),
while genetic properties are hardly recognized in the island context.
This is surprising given the essential role islands have played in the development of evolutionary theories.
